\section{Diffusion Models}
Another class of explicit density estimation generative models is \textbf{diffusion models}. 
Although initially proposed in 2015, they have only recently surged in popularity due to their ability to generate exceptionally high-quality samples. 
Around 2015, Generative Adversarial Networks (GANs) were the state-of-the-art in generative modeling, despite having some serious limitations (as discussed in the previous chapter). 
It was not until 2020 that breakthroughs in architectures and training techniques allowed diffusion models to surpass GANs in sample quality, establishing them as the current state-of-the-art.

Similar to VAEs, diffusion models use neural networks to transform a simple prior distribution (e.g., a standard Gaussian) into a complex data distribution that ideally matches the training data.

At a high level, the fundamental goal of this generative framework involves two simple steps:
\begin{enumerate}
    \item Sample a latent variable $z \sim \mathcal{N}(0, I)$ from a simple distribution;
    \item Pass $z$ through a neural network to generate a sample $\hat{x}$.
\end{enumerate}

While traditional models like GANs might do this in a single pass, diffusion models achieve this transformation iteratively, gradually denoising the latent sample step-by-step. Ultimately, the network is trained to minimize the discrepancy between the generated data and the real data distribution.

Two key questions arise from this overarching framework:
\begin{itemize}
    \item How do we define and structure the latent space distribution?
    \item How do we measure distribution similarity (i.e., what objective function do we use to train the neural network)?
\end{itemize}

\subsection{Idea}
The core idea behind diffusion models is actually quite simple: 
we start with a sample from the data distribution and gradually add noise to it over a series of steps, until it becomes pure noise. 
Then, we train a neural network to reverse this process, learning to denoise the noisy samples step-by-step until we recover the original data sample.

More formally, we define:
\begin{itemize}
    \item \textbf{Forward diffusion process:} starting from a data sample $x_0$ from the data distribution, we define a fixed
    Markov chain that gradually adds noise to the data sample $x_0$ over $T$ steps, resulting in a sequence of noisy samples 
    $x_1, x_2, \ldots, x_T$, where $x_T$ is nearly pure Gaussian noise.

    \item \textbf{Reverse denoising process:} a neural network that learns to reverse the forward diffusion process,
    starting from the noisy sample $x_T$ and iteratively denoising it to recover the original data sample $x_0$.
\end{itemize}

Let's now dive into the details of the forward diffusion process and the reverse denoising process, and how they are implemented in practice.

\subsection{Forward Process}
The forward diffusion process can be defined as a fixed Markov chain that gradually adds Gaussian noise to the data sample $x_0$ over $T$ steps.

\termdef{Markov chain}
{
    A Markov chain is a stochastic process that satisfies the \textbf{Markov property}: 
    the probability distribution of the next state depends only on the current state and not on the sequence of previous states. 
    This property is also known as \textbf{memorylessness}.
}

This implies that the forward process requires no learning, since it is completely specified by a predefined noise schedule.

We can formalize the forward diffusion process as follows:
\begin{equation}
    \label{eq:forward-diffusion}
    q(x_t | x_{t-1}) = \mathcal{N}(\sqrt{{1 - \beta_t}} x_{t-1}, \beta_t I)
\end{equation}
where $\beta_t$ is a variance schedule that controls the amount of noise added at each step. 

Basically, at each step $t$, we take the previous sample $x_{t-1}$, reduce its magnitude by a factor of $\sqrt{1 - \beta_t}$, subsituting it with a random Gaussian noise with variance $\beta_t$.

\subsubsection{A problem}
Suppose we have a data sample $x_0$ and we want to generate its noisy version $x_T$ at step $T$.
We can do this by iteratively applying the forward diffusion (Equation \ref{eq:forward-diffusion}) process $T$ times, starting from $x_0$ and generating $x_1, x_2, \ldots, x_T$.
This would result in:
\[
    q(x_T | x_0) = \prod_{i=1}^{T} q(x_i | x_{i-1})
\]
However, this approach is computationally expensive, especially for large $T$, since we would need to sample from the Gaussian distribution $T$ times, which can be slow and inefficient.

\subsubsection{Reparametrization trick}
Instead, we can use a reparametrization trick to directly sample $x_T$ from $x_0$ in a single step, without having to go through all the intermediate steps.

First, we can rewrite the forward diffusion process in a more convenient form:
\[
    q(x_t | x_{t-1}) = \mathcal{N}(\sqrt{\alpha_t} x_0, (1 - \alpha_t) I)
\]
where $\alpha_t = (1 - \beta_t)$.

By recursively applying this equation, we can derive a closed-form expression for $q(x_T | x_0)$:
\[
    q(x_T | x_0) = \mathcal{N}(\sqrt{\bar{\alpha}_T} x_0, (1 - \bar{\alpha}_T) I)
\]
where $\bar{\alpha}_T = \prod_{t=1}^{T} \alpha_t$.

This makes training and sampling actually feasible, since we can now sample $x_T$ directly from $x_0$ in a single step, without having to go through all the intermediate steps.

\subsection{Reverse Process}
The reverse denoising process is the core of the diffusion model, where we train a neural network to learn the reverse of the forward diffusion process. 
The goal is to start from a noisy sample $x_T$ and iteratively denoise it to recover the original data sample $x_0$.

We could define the reverse process as a Markov chain that learns to reverse the forward diffusion process $p(x_{t-1} | x_t)$.
Applying Bayes' theorem, we can express the reverse process as:
\[
    p(x_{t-1} | x_t) = \frac{p(x_t | x_{t-1}) p(x_{t-1})}{p(x_t)}
\]

But this is intractable in practice, since the marginal distribution $p(x_t)$ would require to know the entire data distribution, at every step $t$, which is not feasible.

Therefore, as previously seen in the VAE chapter, we can train a neural network to approximate the reverse process $p_\theta(x_{t-1} | x_t)$, where $\theta$ are the parameters of the neural network.

By assuming $T \rightarrow \infty$ and $\beta_t \rightarrow 0$, we can show that the reverse process can be approximated as a Gaussian distribution:
\[
    p_\theta(x_{t-1} | x_t) = \mathcal{N}(\mu_\theta(x_t, t), \Sigma_t)
\]
where $\mu_\theta(x_t, t)$ is the mean predicted by the neural network and $\Sigma_t$ is a fixed variance schedule.

Usually, only the mean $\mu_\theta(x_t, t)$ is predicted by the neural network, while the variance $\Sigma_t$ is fixed and known. 
It was later shown that learning the variance $\Sigma_t$ can improve sample quality, but it is not strictly necessary.

\subsubsection{Hierarchical VAEs}

Diffusion models can also be interpreted as a special case of hierarchical variational autoencoders (VAEs). While a standard VAE is formulated with a single latent variable \(z\), a diffusion model introduces a hierarchy of \(T\) latent variables,
\[
    x_1, x_2, \ldots, x_T,
\]
where each variable corresponds to a progressively noisier version of the data sample \(x_0\).

Under this interpretation, diffusion models exhibit several distinctive characteristics:

\begin{itemize}
    \item \textbf{Fixed encoder:} The encoder is not learned but is instead defined by the forward diffusion process, which gradually corrupts the input with Gaussian noise.
    \item \textbf{Hierarchical latent variables:} The latent variables \(x_1, x_2, \ldots, x_T\) form a deep hierarchy, with each variable having the same dimensionality as the original data sample \(x_0\).
    \item \textbf{Shared decoder:} A single neural network is reused at every step of the reverse diffusion process to model the conditional distributions \(p_\theta(x_{t-1}\mid x_t)\), with the timestep \(t\) provided as an additional conditioning input.
\end{itemize}

This interpretation immediatly suggests using the same training objective as in VAEs, which is the Evidence Lower Bound (ELBO).

\subsubsection{Loss}
Following the VAE framework, we can derive the loss function for diffusion models by maximizing the ELBO. The ELBO can be expressed as:

\[
    \mathcal{L} = \mathbb{E}_{q} \left[ 
        \underbrace{D_{KL}(q(x_{T} | x_0) \| p_\theta(x_T))}_{\text{Ensure final Standard Gaussian}} +
        \underbrace{\sum_{t=1}^{T} D_{KL}(q(x_{t-1} | x_t, x_0) \| p_\theta(x_{t-1} | x_t))}_{\text{KL divergence term}} -
        \underbrace{\log p_\theta(x_0 | x_1)}_{\text{Reconstruction term}}
        \right]
\]

Where:
\begin{itemize}
    \item \textbf{First term:} The first term measures the KL divergence between the final noisy sample \(x_T\) and the prior distribution \(p_\theta(x_T)\), which is typically a standard Gaussian. 
    This term ensures that the final latent representation is as close as possible to the prior distribution.
    
    \item \textbf{Second term:} The second term sums the KL divergences between the true posterior \(q(x_{t-1} | x_t, x_0)\) and the learned reverse process \(p_\theta(x_{t-1} | x_t)\) for each timestep \(t\).
    This term encourages the model to learn a reverse process that closely approximates the true posterior, effectively guiding the denoising process.

    \item \textbf{Third term:} The third term is the reconstruction term, which measures how well the model can reconstruct the original data sample \(x_0\) from the first noisy sample \(x_1\).
    This term ensures that the model learns to generate samples that are faithful to the original data distribution.
\end{itemize}

From the ELBO, we can derive a simplified loss function that is more practical for training diffusion models,
that focuses on predicting the noise \(\epsilon\) added to the original data sample \(x_0\) at each timestep \(t\).
\begin{equation}
    \| \epsilon - \epsilon_\theta(x_t, t) \|^2
\end{equation}
Where \(\epsilon_\theta(x_t, t)\) is the noise predicted by the neural network at timestep \(t\).

By applying some further simplifications and reparametrizations, we can express the final loss function as:
\[
    \mathcal{L} = \mathbb{E}_{x_0, \epsilon, t} \left[ \| \epsilon - \underbrace{\epsilon_\theta(\sqrt{\bar{\alpha}_t} x_0 + \sqrt{1 - \bar{\alpha}_t} \epsilon, t)}_{\text{Predicted noise}} \|^2 \right]
\]
Where, at a high level, we are basically picking a random timestep \(t\), a random noise \(\epsilon\), and a random data sample \(x_0\). 
In the forward process, we generate a noisy sample \(x_t\) by adding the noise \(\epsilon\) to the original data sample \(x_0\) according to the forward diffusion process. 
Then, we feed this noisy sample \(x_t\) into the neural network to predict the noise \(\epsilon_\theta(x_t, t)\), and we compute the mean squared error between the predicted noise and the true noise \(\epsilon\).

\subsubsection{Sampling}
At sampling time, we start from a sample \(x_T\) drawn from the prior distribution (typically a standard Gaussian) and iteratively apply the learned reverse process, which is defined as:

\begin{equation}
    x_{t-1} = \frac{1}{\sqrt{\alpha_t}} \left( x_t - \frac{1 - \alpha_t}{\sqrt{1 - \bar{\alpha}_t}} \epsilon_\theta(x_t, t) \right) + \sigma_t z
\end{equation}
This formulation allow to re-weight the predicted noise \(\epsilon_\theta(x_t, t)\) based on the current timestep \(t\).
In fact, early in the reverse process, when \(t\) is large, the updates are conservative, whereas later in the process, when \(t\) is small, the updates become more aggressive, allowing the model to refine the generated sample as it approaches the original data distribution.

\subsection{The Generative Trilemma}
The generative models landscape is often characterized by a trade-off between three key aspects: sample quality, diversity, and inference speed. 
This trade-off is commonly referred to as the \textbf{Generative Trilemma}.

In particular:
\begin{itemize}
    \item \textbf{GANs:} known for their high sample quality and fast inference speed, but they often suffer from mode collapse, leading to low diversity in generated samples.
    \item \textbf{VAEs:} known for their high diversity and fast inference, but they often produce samples of lower quality due to the limitations of the variational approximation.
    \item \textbf{Diffusion Models:} known for their high sample quality and high diversity, but they are typically slower in inference due to the iterative nature of the denoising process.
\end{itemize}

\subsection{Architectures}

Again, similar to VAEs, diffusion models can be implemented using a variety of neural network architectures. 
We can distinguish between \textbf{unconditional} and \textbf{conditional} diffusion models, depending on whether the model is conditioned on additional information (e.g., class labels, text prompts, etc.) or not.

\subsubsection{Unconditional Models}
\paragraph{U-Net}
The most common architecture for image diffusion models is the \textbf{U-Net}, a convolutional neural network originally introduced for image segmentation. 
It consists of an \textbf{encoder-decoder} structure with skip connections that preserve spatial information across multiple resolutions, enabling the reconstruction of fine-grained image details.

Many diffusion models also incorporate \textbf{self-attention} layers within the U-Net to capture long-range dependencies between distant regions of the image, improving the quality and global coherence of the generated samples.

Since the denoising network must condition its predictions on the current diffusion timestep \(t\), the timestep is embedded into a continuous vector representation before being provided to the network. 
A common approach is to use \textbf{sinusoidal embeddings} (also known as Fourier features), which encode \(t\) using sine and cosine functions at multiple frequencies. 

\paragraph{Transformers - DiT}
Later, in 2021, researchers proposed using \textbf{transformers} as the backbone architecture for diffusion models, leading to the development of \textbf{DiT} (Diffusion Transformer).
This approach, although more computationally expensive, shows that transformers can effectively model the denoising process, better than convolutional architectures.

\subsubsection{Consitionla Models}
