\section{Expanding CNN Beyond Classification}
In the previous chapter, we have seen how Convolutional Neural Networks (CNNs) can be used for image classification tasks. 
However, CNNs can be applied to a wide range of other tasks beyond classification, such as \textbf{object detection}, 
\textbf{semantic segmentation}, and \textbf{captioning}, among others.
In this chapter, we will explore some of these applications and how CNNs can be adapted to solve them.

\subsection{Object Detection}
Object detection is the task of identifying and localizing objects within an image. It can be seen as a combination of
classification (what object is present) and regression (where is the object located).
The task input would be an image, and the output would be a set of bounding boxes around the detected objects, along with their corresponding class labels.

This task is more complex than image classification, as it requires the model to not only recognize the presence of objects but also to accurately localize them within the image.
This makes also the training process more challenging, as several issues can arise, such as:
\begin{itemize}
    \item \textbf{Data scarcity}: Object detection datasets are often smaller than classification datasets. Also,
    being the task more complex, it requires more data to train a model that can generalize well.
    \item \textbf{Poor localization}: the deep CNN proposed for classification may not be able to localize the objects accurately, 
    because of their large receptive field and the pooling layers that reduce the spatial resolution of the feature maps.
\end{itemize}

To address these issues, several architectures have been proposed, such as \textbf{R-CNN}, \textbf{Fast R-CNN}, and \textbf{YOLO} (You Only Look Once), among others.

\subsubsection{Before CNNs: DPM}
Before AlexNet and the rise of CNNs, one of the most popular approaches for object detection was the Deformable Part Model (DPM).
DPM is a method that models objects as a collection of parts, where each part can be described by a set of features. 
The model learns to detect these parts and their spatial relationships to identify the presence of an object in an image.

This approach was quite effective for its time, but reached a plateau in performance already in 2010, before the advent of deep learning.

\subsubsection{Selective Search}
One of the key challenges in object detection is to efficiently search for objects in an image.
Selective Search is a method that generates a set of candidate regions (also called proposals) in an image
that are likely to contain objects. 
Given the proposals, the process of object detection can be simplified to a classification task, where the model only needs to classify the proposals as containing an object or not,
which is much easier than searching for objects in the entire image.

To detect objects at any scale, Selective Search uses a hierarchical grouping of superpixels, which are small regions of an image that 
have similar features (such as color, texture, or intensity).
The algorithm starts by generating all the possible candidate objects. Similar regions
are then merged together based on their similarity. 
Once the merging process is complete, the algorithm uses the resulting regions as proposals for object detection.

\subsubsection{R-CNN}
R-CNN (Region-based Convolutional Neural Networks) is one of the first architectures that successfully applied CNNs to object detection. 
The R-CNN architecture consists of three main steps:
\begin{itemize}
    \item \textbf{Region Proposal}: The first step is to generate a set of candidate regions (proposals) in the image using Selective Search.
    \item \textbf{Feature Extraction}: The second step is to extract features from each proposal using a CNN.
    \item \textbf{Classification}: The final step is to classify each proposal as containing an object or not, and to assign a class label to the detected objects.
    This was typically done using a Support Vector Machine (SVM) classifier, which was trained on the features extracted from the proposals.
\end{itemize}

This approach was a significant improvement over previous methods, but had a few limitations, such as:
\begin{itemize}
    \item \textbf{NOT an end-to-end architecture}: The R-CNN architecture is not an end-to-end architecture, as it requires separate training for the region proposal, feature extraction, and classification steps.
    Deep learning models are known to perform better when trained end-to-end, as they can learn to optimize the entire process jointly.
    \item \textbf{Image resizing}: Each proposal must be resized to a fixed size (e.g., 224x224) before being fed into the CNN for feature extraction.
    This can lead to a loss of information, especially for small objects, and can also introduce distortions in the aspect ratio of the proposals.
    \item \textbf{Computationally expensive}: The R-CNN architecture is computationally expensive, as it requires running the CNN for each proposal, 
    which can be in the order of thousands per image (takes around 47 seconds per image on a GPU).
\end{itemize}

\subsubsection{Fast R-CNN}
\subsection{Semantic Segmentation}