\subsection{From Logistic Regression to Feedforward NN}
To fit a model to some data distribution, a simple and common approach is to use \textbf{Maximum Likelihood Estimation} (MLE).

\subsubsection{Maximum Likelihood Estimation}
In MLE, we want to estimate the parameters of an assumed probability distribution, given some observed data. For example, if we assume that the data is generated from a Gaussian distribution, we can estimate the mean and variance of that distribution using MLE.

More formally, given a dataset $\boldsymbol{X} = \{\boldsymbol{x}_1, \boldsymbol{x}_2, \ldots, \boldsymbol{x}_N\}$ and a known probability distribution $p_{\text{model}}(\boldsymbol{x}|\boldsymbol{\theta})$
we want to find the parameters $\boldsymbol{\theta}$ that better explain the data.
\begin{equation}
    \label{MLE}
    \boldsymbol{\theta}^* = \arg\max_{\boldsymbol{\theta}} p_{\text{model}}(\boldsymbol{X}|\boldsymbol{\theta})
\end{equation}
Assuming that the data points are independent and identically distributed (i.i.d.), we can express the likelihood as a product of individual probabilities:
\begin{equation*}
    p_{\text{model}}(\boldsymbol{X}|\boldsymbol{\theta}) = \prod_{i=1}^{N} p_{\text{model}}(\boldsymbol{x}_i|\boldsymbol{\theta})
\end{equation*}
To simplify the optimization process, we often take the logarithm of the likelihood function, which transforms the product into a sum:
\begin{equation*}
    \log p_{\text{model}}(\boldsymbol{X}|\boldsymbol{\theta}) = \sum_{i=1}^{N} \log p_{\text{model}}(\boldsymbol{x}_i|\boldsymbol{\theta})
\end{equation*}
We can show that this is equivalent to considering the expected value of the log-likelihood over the empirical (observed) data distribution:
\begin{equation*}
    \log p_{\text{model}}(\boldsymbol{X}|\boldsymbol{\theta}) = N \cdot \mathbb{E}_{\boldsymbol{x} \sim p_{\text{data}}}[\log p_{\text{model}}(\boldsymbol{x}|\boldsymbol{\theta})]
\end{equation*}
where $p_{\text{data}}$ is the empirical distribution of the observed data and $\mathbb{E}_{\boldsymbol{x} \sim p_{\text{data}}}$ denotes the expected value sampled from this distribution.

To find the optimal parameters $\boldsymbol{\theta}^*$, we usually differentiate the log-likelihood with respect to $\boldsymbol{\theta}$ and set it
to zero. We can then resolve the resulting equation to find the optimal parameters.

The equation \ref{MLE} can also be extended to the case of conditional probability distributions ($p_{\text{model}}(y|\boldsymbol{x}, \boldsymbol{\theta})$) in order to predict a target variable $y$ given an input variable $\boldsymbol{x}$:
\begin{equation*}
    \boldsymbol{\theta}^* = \arg\max_{\boldsymbol{\theta}} p_{\text{model}}(y|\boldsymbol{x}, \boldsymbol{\theta})
\end{equation*}
This is the basis for most supervised learning algorithms, including Linear Regression and Logistic Regression, which simply assume different forms for the conditional probability distribution.
For example, in Logistic Regression, we assume that the conditional probability distribution is given by the logistic function:
\begin{equation}
    \label{Logistic Regression}
    p_{\text{model}}(y=1|\boldsymbol{x}, \boldsymbol{\theta}) = \sigma(\boldsymbol{\theta}^T \boldsymbol{x}) = \frac{1}{1 + e^{-\boldsymbol{\theta}^T \boldsymbol{x}}}
\end{equation}

\subsubsection{Logistic Regression}
Logistic Regression is a linear model used for binary classification tasks.
\begin{comment}
    #TODO: extend here??
\end{comment}
Given the equation \ref{Logistic Regression}, we can observe that the can be seen as a simple Neural Network with a single layer and a sigmoid activation function.